\section{Conclusion}
\label{sec:conclusion}

We introduced a small, compositional static analysis that makes path-insensitive precision loss explicit via a graded algebra of types. Base types and $\Any$ carry a nonnegative grade; same-grade heterogeneous joins are conservative, and grades increase only when the analysis commits to a base after imprecision ($\Any^{i}\!\le\!X^{\,i+1}$). A cost-based interpretation selection for overloaded operators ensures that the abstract interpreter assigns principal (most precise) types: it prefers the interpretation requiring fewest coercions, joining only on ties. We instantiated this domain in a standard abstract interpreter for a \textsf{while}-language, proved monotonicity and soundness (as a may-analysis), and explained termination either by a simple grade-capping widening or by a finite ghost-set product domain. The grade and its associated culprit-set provide actionable diagnostics: they localize where assumptions were forced and bound how much imprecision a value carries.

The approach is deliberately cast-free and leaves operational semantics unchanged. The relationship to gradual typing is not merely conceptual but implemented and demonstrated: the abstract interpreter is parameterized over a single coercion function, and substituting the grade successor ($n\mapsto\infty$ instead of $n\mapsto n{+}1$) reproduces standard gradual typing exactly---confirming that the entire behavioral difference is one function. Extensions---grade-parametric function signatures, grade-as-effect grades, shape-preserving joins, SSA $\phi$-site instrumentation, and the realized gradual mode---fit without disturbing the core algebra.

Limitations are clear: the analysis is path-insensitive, does not guarantee absence of runtime type errors, and we have not attempted an empirical evaluation here. Future work in \S\ref{sec:future} outlines next steps (graded semantics, precision recovery, mechanization). The core contribution is a minimal algebra and analysis that expose, rather than collapse, the structure of precision loss-yielding principled and explainable static diagnostics.
